---
title: "Validación externa de KFRE para pronostico de falla renal en asegurados de EsSalud"
author: 
  - name: Percy Soto-Becerra
    url: https://github.com/psotob91
date: last-modified
lang: es
format: 
  html:
    self-contained: true
    number-sections: true
    number-depth: 3
    anchor-sections: true
    code-fold: true
    code-summary: "Mostrar código"
    code-overflow: wrap
    code-tools: true
    code-link: true
    # code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    theme: 
      light: journal
      dark: darkly
    highlight-style: espresso
    mainfont: Source Sans Pro
    html-math-method: katex
    toc: true
editor: visual

execute: 
  warning: false
  message: false
  error: false
  echo: false
  column: page
---

```{r}
#| label: load-packages
#| include: false

# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_unload("all")

if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install(version = "3.15")
}

if (!require("survcomp")) 
    BiocManager::install("survcomp")

pacman::p_load(
  here, 
  skimr, 
  survival,
  rms,
  cmprsk,
  riskRegression,
  mstate,
  pseudo,
  pec,
  riskRegression,
  plotrix,
  knitr,
  splines,
  kableExtra,
  flextable,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot, 
  patchwork,
  survival, 
  cmprsk, 
  survminer, 
  ggsci, 
  cowplot, 
  scales, 
  patchwork, 
  labelled, 
  glue, 
  dcurves, 
  broom, 
  downlit, 
  xml2
)

data <- readRDS(here::here("Data/Derived/data_derived_v2.rds")) 

# Data Stages 3-4
data %>%  
  filter(ckd_stage == "Stages 3-4") -> dataA

# Data Stages 3b-4
data %>%  
  filter(ckd_stage2 == "Stages 3b-4") -> dataB
```

## Análisis descriptivo

En total, `r nrow(dataA)` pacientes fueron incluidos en el análisis del grupo con ERC estadíos 3a-4 y `r nrow(dataB)` en el grupo ERC con estadíos 3b-4. Todos los pacientes tuvieron datos del desenlace. El número de eventos fue mayor a 100 en todas las poblaciones y desenlaces, a excepción de la falla renal a 2 años en la población con ERC estadíos 3b-4 (n = 88), por lo que los estimados en este grupo deben interpretarse con mucha más cautela. Específicamente, en el grupo ERC-3a-3b-4, `r table(dataA$eventd2y)[2]` (`r round(prop.table(table(dataA$eventd2y))[2] * 100, 1)`%) pacientes desarrollaron falla renal en 2 años y `r table(dataA$eventd5y)[2]` (`r round(prop.table(table(dataA$eventd5y))[2] * 100, 1)`%) en 5 años. Muchos más pacientes murieron sin experimentar falla renal: `r table(dataA$eventd2y)[3]` (`r round(prop.table(table(dataA$eventd2y))[3] * 100, 1)`%) en 2 años y `r table(dataA$eventd5y)[3]` (`r round(prop.table(table(dataA$eventd5y))[3] * 100, 1)`%) en 5 años.

Respecto al grupo restringido a estadios ERC-3b-4, `r table(dataB$eventd2y)[2]` (`r round(prop.table(table(dataB$eventd2y))[2] * 100, 1)`%) pacientes desarrollaron falla renal en 2 años y `r table(dataB$eventd5y)[2]` (`r round(prop.table(table(dataB$eventd5y))[2] * 100, 1)`%) en 5 años. Del mismo modo, muchos más pacientes murieron sin experimentar falla renal: `r table(dataB$eventd2y)[3]` (`r round(prop.table(table(dataB$eventd2y))[3] * 100, 1)`%) en 2 años y `r table(dataB$eventd5y)[3]` (`r round(prop.table(table(dataB$eventd5y))[3] * 100, 1)`%) en 5 años.

Todos los pacientes fueron censurados administrativamente el 31 de diciembre de 2019. La mediana de tiempo de observación fue de `r round(median(dataA$time, na.rm = TRUE), 1)` años el seguimiento máximo fue de `r round(max(dataA$time, na.rm = TRUE), 1)` años en el grupo ERC 3a-4. La mediana y máximo de tiempos de observación fue similar en el grupo con ERC 3b-4 con `r round(median(dataA$time, na.rm = TRUE), 1)` y `r round(max(dataA$time, na.rm = TRUE), 1)` años, respectivamente. En total, `r table(dataA$eventd2y)[1]` pacientes (`r round(prop.table(table(dataA$eventd2y))[1] * 100, 1)`%) fueron censurados administrativamente en 2 años y `r table(dataA$eventd5y)[1]` pacientes (`r round(prop.table(table(dataA$eventd5y))[1] * 100, 1)`%) en 5 años.

### Tabla 1

La [Tabla 1] describe a la población de estudio. Las [Tabla S1](#sec-tableS1) detalla las caracterísitcas de la población de estudio en estadios 3a-3b-4 según la ocurrencia del desenlace de interés falla renal a 2 y 5 años. El número de eventos de interés a 2 años fue de La [Tabla S2](#sec-tableS2) muestra esto mismo para la población 3b-4. La [Tabla S3](#sec-tableS3) describe las características según estadíso 3a, 3b y 4 por separado. Con esta estratificación, los números de casos de falla renal a 2 años son muy bajos para las subpoblaciones con estadíos 3a (n = `r table(data$eventd2y, data$grf_cat)[2, 1]`), 3b (n = `r table(data$eventd2y, data$grf_cat)[2, 2]`) y 4 (n = `r table(data$eventd2y, data$grf_cat)[2, 3]`). Del mismo modo, los números de casos a 5 años fueron muy bajos para las subpoblaciones con estadíos 3a (n = `r table(data$eventd5y, data$grf_cat)[2, 1]`), 3b (n = `r table(data$eventd5y, data$grf_cat)[2, 2]`) y 4 (n = `r table(data$eventd5y, data$grf_cat)[2, 3]`). Por tal motivo, no fue confiable realizar la evaluación del rendimiento predictivo en estos subgrupos.

Por último, la [Tabla S4](#sec-tableS4) describe la distribución de la población de estudio en cada centro de salud de Lima.

```{r}
#| label: table1

data %>%  
  filter(ckd_stage == "Stages 3-4") %>% 
  dplyr::select(total, sex, age, acr_cat, grf_cat, ckd_class, crea, eGFR_ckdepi, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, death2y, eventd2ylab, death5y, eventd5ylab) %>% 
  gtsummary::tbl_summary(
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25} - {p75})", 
    "{min} - {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  # add_p() %>% 
  bold_labels() -> tab1a

data %>%  
  filter(ckd_stage2 == "Stages 3b-4") %>% 
  mutate(
    grf_cat = droplevels(grf_cat), 
    ckd_class = droplevels(ckd_class)
  ) %>% 
  set_variable_labels(
    grf_cat = "GFR categories", 
    ckd_class = "CKD KDIGO classification"
  ) %>% 
  dplyr::select(total, sex, age, acr_cat, grf_cat, ckd_class, crea, eGFR_ckdepi, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, death2y, eventd2ylab, death5y, 
                eventd5ylab) %>% 
  gtsummary::tbl_summary(
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25} - {p75})", 
    "{min} - {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  # add_p() %>% 
  bold_labels() -> tab1b

tbl_merge(list(tab1a, tab1b), 
          tab_spanner = c("**CKD Stages 3a-3b-4**", "**CKD Stages 3b-4**")) %>% 
  modify_caption("Table 1. Baseline characteristics of the study population according CKD Stages") %>% 
  bold_labels() -> tab1

tab1 %>% 
  as_flex_table() %>% 
  flextable::save_as_docx(path = "Tables/Table1.docx")

tab1 %>% 
  gtsummary::as_kable_extra() %>%
  kableExtra::kable_styling("striped")
```

### Curvas de función de incidencia cumulada

```{r}
#| label: table2-code

# Seleccion del grupo----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

# timepoints(cif, times = 1:5) #< Da lo mismo que usar crprep

vdata.w1 <- crprep(
  Tstop = "time5y",
  status = "eventd5y",
  trans = c(1, 2),
  id = "id",
  keep = c("age", "male", "eGFR_mdrd", "acr", "risk2y", "risk5y"),
  data = data_filt
)

vdata.w1 <- vdata.w1 %>% filter(failcode == 1)
vdata.w2 <- vdata.w1 %>% filter(failcode == 2)

mfit_vdata1 <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = vdata.w1, 
  weights = weight.cens
)

smfit_vdata1 <- summary(mfit_vdata1, times = c(1, 2, 3, 4, 5))

res_ci_stg12 <- cbind(
  100 * (1 - smfit_vdata1$surv),
  100 * (1 - smfit_vdata1$upper),
  100 * (1 - smfit_vdata1$lower)
)

res_ci_stg12 <- round(res_ci_stg12, 2)

rownames(res_ci_stg12) <- c(
  "1-year", "2-year",
  "3-year", "4-year",
  "5-year"
)

colnames(res_ci_stg12) <- c(
  "Estimate", "Lower .95",
  "Upper .95"
)

# Seleccion del grupo----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata.w1 <- crprep(
  Tstop = "time5y",
  status = "eventd5y",
  trans = c(1, 2),
  id = "id",
  keep = c("age", "male", "eGFR_mdrd", "acr", "risk2y", "risk5y"),
  data = data_filt
)

vdata.w1 <- vdata.w1 %>% filter(failcode == 1)
vdata.w2 <- vdata.w1 %>% filter(failcode == 2)

mfit_vdata1 <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = vdata.w1, 
  weights = weight.cens
)

smfit_vdata1 <- summary(mfit_vdata1, times = c(1, 2, 3, 4, 5))

res_ci_stg35 <- cbind(
  100 * (1 - smfit_vdata1$surv),
  100 * (1 - smfit_vdata1$upper),
  100 * (1 - smfit_vdata1$lower)
)

res_ci_stg35  <- round(res_ci_stg35, 2)

rownames(res_ci_stg35) <- c(
  "1-year", "2-year",
  "3-year", "4-year",
  "5-year"
)

colnames(res_ci_stg35) <- c(
  "Estimate", "Lower .95",
  "Upper .95"
)

res_ci <- cbind(res_ci_stg12, res_ci_stg35)

res_ci %>% 
  as_tibble(rownames = "Year") %>% 
  mutate(
    est1 = str_glue("{Estimate}%"), 
    est1.ci = str_glue(" ({`Lower .95`}% to {`Upper .95`}%)"), 
    est2 = str_glue("{V4}%"), 
    est2.ci = str_glue(" ({V5}% to {V6}%)")
  ) %>% 
  select(Year, est1, est1.ci, est2, est2.ci) %>% 
  mutate(across(everything(), ~as.character(.x))) -> res_ci2 

res_ci2 %>% 
  flextable() %>% 
  add_header_row(values = c("Year", "CKD Stages 3a-3b-4", "CKD Stages 3b-4"), colwidths = c(1, 2, 2)) %>% 
  set_header_labels(
    est1 = "%", 
    est1.ci = "95% CI", 
    est2 = "%", 
    est2.ci = "95% CI" 
  ) %>% 
  merge_v(j = 1, part = "header") %>% 
  set_caption("Table 2. Cumulative incidence of kidney failure") %>% 
  theme_box() %>% 
  autofit() -> tab_cif

tab_cif %>% 
  save_as_docx(path = "Tables/Table2.docx")
```

La incidencia acumulada estimada de falla renal a 2 años y a 5 años fue de `r res_ci2$est1[2]` `r res_ci2$est1.ci[2]` y `r res_ci2$est1[5]` `r res_ci2$est1.ci[5]` en pacientes con ERC estadíos 3a, 3b o 4, respectivamente. En pacientes con ERC 3b-4, las incidicencias acumuladas a 2 y 5 años fueron de `r res_ci2$est2[2]` `r res_ci2$est2.ci[2]` y `r res_ci2$est2[5]` `r res_ci2$est2.ci[5]`, respectivamente. La [Tabla S6](#sec-table6) detalla las incidencias acumuladas de falla renal 1 a 5 años en ambas cohortes.

La @fig-cif muestra las curvas de incidencia acumulada de falla renal y muerte pre-falla renal en los pacientes del estudio.

```{r}
#| label: cif-1
#| output: false

# Seleccion del grupo----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

cuminc(ftime = data_filt$time5y, 
       fstatus = data_filt$eventd5ylab, 
       cencode = "Alive w/o Kidney Failure") -> cif

ciplotdat <- 
  cif %>% 
  list_modify("Tests" = NULL) %>% 
  map_df(`[`, c("time", "est", "var"), .id = "id") %>% 
  mutate(id = recode(
    id, 
    "1 Death w/o Kidney Failure" = "Death w/o Kidney Failure", 
    "1 Kidney Failure" = "Kidney Failure"), 
    ll = est - 1.96 * sqrt(var), 
    ul = est + 1.96 * sqrt(var)
    ) %>% 
  rename(
    event = id
  )

ciplotdat %>% 
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul, fill = event), 
              alpha = 0.25, linetype = 0) + 
  geom_step(lwd = 0.5, aes(color = event)) +
  theme_survminer() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
  labs(x = "Years", 
       y = "Cumulative incidence",
       title = "A CKD Stages 3a-3b-4") + 
  theme(legend.position = "top",
        legend.title = element_blank(), 
        legend.background = element_rect(fill = "white"), 
        legend.key.size = unit(0.2, "cm")) -> g1

ciplotdat %>% 
  filter(event == "Kidney Failure") %>%
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "#00BFC4", 
              alpha = 0.1, linetype = 0) + 
  geom_step(lwd = 0.5, color = "#00BFC4") +
  theme_survminer() +
  ylim(c(0, 0.025)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.037)) + 
  labs(x = "", 
       y = "",
       title = "") -> g2

kf_fit <- survfit(
  Surv(time5y, ifelse(eventd5y != 0, 1, 0)) ~ 1, 
  data = data_filt
)

num <- ggsurvplot(
  fit = kf_fit, 
  risk.table = "nrisk_cumevents", 
  risk.table.y.text = FALSE,
  risk.table.y.text.col = FALSE, 
  tables.y.text = FALSE, 
  tables.y.text.col = FALSE, 
  ylab = "Years",
  risk.table.fontsize = 3.2,
  tables.theme = theme_survminer(font.main = 10)
  )

cowplot::plot_grid(
  g1, 
  num$table + theme_cleantable(), 
  nrow = 2, 
  rel_heights = c(4, 1), 
  align = "v", 
  axis = "b"
  ) -> g3
  
g3 + inset_element(g2, 0.15, 0.43, 1, 0.856,  align_to = 'full',  
                   ignore_tag = TRUE) -> plot_cif_mh
 
# Seleccion del grupo----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

cuminc(ftime = data_filt$time5y, 
       fstatus = data_filt$eventd5ylab, 
       cencode = "Alive w/o Kidney Failure") -> cif

ciplotdat <- 
  cif %>% 
  list_modify("Tests" = NULL) %>% 
  map_df(`[`, c("time", "est", "var"), .id = "id") %>% 
  mutate(id = recode(
    id, 
    "1 Death w/o Kidney Failure" = "Death w/o Kidney Failure", 
    "1 Kidney Failure" = "Kidney Failure"), 
    ll = est - 1.96 * sqrt(var), 
    ul = est + 1.96 * sqrt(var)
    ) %>% 
  rename(
    event = id
  )

ciplotdat %>% 
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul, fill = event), 
              alpha = 0.25, linetype = 0) + 
  geom_step(lwd = 0.5, aes(color = event)) +
  theme_survminer() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
  labs(x = "Years", 
       y = "Cumulative incidence",
       title = "B CKD Stages 3b-4") + 
  theme(legend.position = "top",
        legend.title = element_blank(), 
        legend.background = element_rect(fill = "white"), 
        legend.key.size = unit(0.2, "cm")) -> g1

ciplotdat %>% 
  filter(event == "Kidney Failure") %>%
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "#00BFC4", 
              alpha = 0.1, linetype = 0) + 
  geom_step(lwd = 0.5, color = "#00BFC4") +
  theme_survminer() +
  ylim(c(0, 0.15)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.19)) + 
  labs(x = "", 
       y = "",
       title = "") -> g2

kf_fit <- survfit(
  Surv(time5y, ifelse(eventd5y != 0, 1, 0)) ~ 1, 
  data = data_filt
)

num <- ggsurvplot(
  fit = kf_fit, 
  risk.table = "nrisk_cumevents", 
  risk.table.y.text = FALSE,
  risk.table.y.text.col = FALSE, 
  tables.y.text = FALSE, 
  tables.y.text.col = FALSE, 
  ylab = "Years",
  risk.table.fontsize = 3.2,
  tables.theme = theme_survminer(font.main = 10)
  )

cowplot::plot_grid(
  g1, 
  num$table + theme_cleantable(), 
  nrow = 2, 
  rel_heights = c(4, 1), 
  align = "v", 
  axis = "b"
  ) -> g3

g3 + inset_element(g2, 0.15, 0.51, 1, 0.856,  align_to = 'full',  
                   ignore_tag = TRUE) -> plot_cif_vh

(plot_cif_mh / plot_cif_vh) + 
  plot_annotation(tag_levels = 'A') -> plot_cif

ggsave(filename = "Plot_CIF.png", 
      plot = plot_cif, 
      device = "png", 
      path = "Figures", 
      dpi = 300, 
      scale = 2, 
      width = 8.5,
      height = 14, 
      units = "cm", 
      bg = "white")
```

```{r}
#| label: fig-cif
#| fig.width: 7
#| fig.height: 14
#| fig-cap: "Cumulative incidence curves of the observed outcome probabilities in the population study for kidney failure or deathd without kidney failure"

plot_cif
```

```{r}
#| label: cif-2
#| output: false

# Seleccion del grupo----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

cuminc(ftime = data_filt$time5y, 
       fstatus = data_filt$eventd5ylab, 
       cencode = "Alive w/o Kidney Failure") -> cif

ciplotdat <- 
  cif %>% 
  list_modify("Tests" = NULL) %>% 
  map_df(`[`, c("time", "est", "var"), .id = "id") %>% 
  mutate(id = recode(
    id, 
    "1 Death w/o Kidney Failure" = "Muerte sin Falla Renal", 
    "1 Kidney Failure" = "Falla Renal"), 
    ll = est - 1.96 * sqrt(var), 
    ul = est + 1.96 * sqrt(var)
    ) %>% 
  rename(
    event = id
  )

ciplotdat %>% 
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul, fill = event), 
              alpha = 0.25, linetype = 0) + 
  geom_step(lwd = 0.5, aes(color = event)) +
  theme_survminer() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
  labs(x = "Años", 
       y = "Incidencia Acumulada",
       title = "A ERC Estadios 3a-3b-4") + 
  theme(legend.position = "top",
        legend.title = element_blank(), 
        legend.background = element_rect(fill = "white"), 
        legend.key.size = unit(0.2, "cm")) -> g1

ciplotdat %>% 
  filter(event == "Falla Renal") %>%
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "#00BFC4", 
              alpha = 0.1, linetype = 0) + 
  geom_step(lwd = 0.5, color = "#00BFC4") +
  theme_survminer() +
  ylim(c(0, 0.025)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.04)) + 
  labs(x = "", 
       y = "",
       title = "") -> g2

kf_fit <- survfit(
  Surv(time5y, ifelse(eventd5y != 0, 1, 0)) ~ 1, 
  data = data_filt
)

num <- ggsurvplot(
  fit = kf_fit, 
  risk.table = "nrisk_cumevents", 
  risk.table.y.text = FALSE,
  risk.table.y.text.col = FALSE, 
  risk.table.title = "Número en riesgo (Número de eventos)", 
  tables.y.text = FALSE, 
  tables.y.text.col = FALSE, 
  ylab = "Años",
  risk.table.fontsize = 3.2,
  tables.theme = theme_survminer(font.main = 10)
  )

cowplot::plot_grid(
  g1, 
  num$table + theme_cleantable(), 
  nrow = 2, 
  rel_heights = c(4, 1), 
  align = "v", 
  axis = "b"
  ) -> g3
  
g3 + inset_element(g2, 0.15, 0.43, 1, 0.856,  align_to = 'full',  
                   ignore_tag = TRUE) -> plot_cif_mh

# Seleccion del grupo----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

cuminc(ftime = data_filt$time5y, 
       fstatus = data_filt$eventd5ylab, 
       cencode = "Alive w/o Kidney Failure") -> cif

ciplotdat <- 
  cif %>% 
  list_modify("Tests" = NULL) %>% 
  map_df(`[`, c("time", "est", "var"), .id = "id") %>% 
  mutate(id = recode(
    id, 
    "1 Death w/o Kidney Failure" = "Muerte sin Falla Renal", 
    "1 Kidney Failure" = "Falla Renal"), 
    ll = est - 1.96 * sqrt(var), 
    ul = est + 1.96 * sqrt(var)
    ) %>% 
  rename(
    event = id
  )

ciplotdat %>% 
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul, fill = event), 
              alpha = 0.25, linetype = 0) + 
  geom_step(lwd = 0.5, aes(color = event)) +
  theme_survminer() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
  labs(x = "Años", 
       y = "Incidencia Acumulada",
       title = "B ERC Estadios 3b-4") + 
  theme(legend.position = "top",
        legend.title = element_blank(), 
        legend.background = element_rect(fill = "white"), 
        legend.key.size = unit(0.2, "cm")) -> g1

ciplotdat %>% 
  filter(event == "Falla Renal") %>%
  ggplot(aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "#00BFC4", 
              alpha = 0.1, linetype = 0) + 
  geom_step(lwd = 0.5, color = "#00BFC4") +
  theme_survminer() +
  ylim(c(0, 0.15)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.19)) + 
  labs(x = "", 
       y = "",
       title = "") -> g2

kf_fit <- survfit(
  Surv(time5y, ifelse(eventd5y != 0, 1, 0)) ~ 1, 
  data = data_filt
)

num <- ggsurvplot(
  fit = kf_fit, 
  risk.table = "nrisk_cumevents", 
  risk.table.y.text = FALSE,
  risk.table.y.text.col = FALSE, 
  risk.table.title = "Número en riesgo (Número de eventos)", 
  tables.y.text = FALSE, 
  tables.y.text.col = FALSE, 
  ylab = "Años",
  risk.table.fontsize = 3.2,
  tables.theme = theme_survminer(font.main = 10)
  )

cowplot::plot_grid(
  g1, 
  num$table + theme_cleantable(), 
  nrow = 2, 
  rel_heights = c(4, 1), 
  align = "v", 
  axis = "b"
  ) -> g3

g3 + inset_element(g2, 0.15, 0.51, 1, 0.856,  align_to = 'full',  
                   ignore_tag = TRUE) -> plot_cif_vh

(plot_cif_mh / plot_cif_vh) + 
  plot_annotation(tag_levels = 'A') -> plot_cif

ggsave(filename = "Plot_CIF_Spanish.png", 
      plot = plot_cif, 
      device = "png", 
      path = "Figures", 
      dpi = 300, 
      scale = 2, 
      width = 10,
      height = 14, 
      units = "cm", 
      bg = "white")
```

## Evaluación de rendimiento predictivo:

A continuación, reportaremos los resultados de evaluación del rendimiento predictivo del modelo KFRE para dos horizontes de predicción: 2 años y 5 años. Evaluamos tres aspectos del rendimiento predictivo:

1)  Calibración del modelo

2)  Discriminación del modelo

3)  Rendimiento global

::: {.callout-note collapse="true"}
## Nota

De acuerdo con las recomendaciones dadas por la guía [TRIPOD](https://www.acpjournals.org/doi/full/10.7326/M14-0698?rfr_dat=cr_pub++0pubmed&url_ver=Z39.88-2003&rfr_id=ori%3Arid%3Acrossref.org) (Transparent Reporting of multivariable prediction model for Individual Prognosis or Diagnosis), la validación (interna y externa) de todo modelo de predicción (diagnóstico o pronóstico) debería reportar al menos dos aspectos claves de su rendimiento predicitivo: calibración y discriminación. Adicionalmente, se puede optar por medidas de rendimiento global.
:::

### Calibración del modelo

::: {.callout-note collapse="true"}
## Nota

Evaluamos calibración mediante:

-   **Gráficos de calibración**:
    -   Curvas suavizadas (mediante regresión local) y cinturón de calibración (intervalos de confianza al 95%) de proporción observada versus probabilidad predicha por modelo.
    -   La proporción observada puede ser estimada por varios métodos que tienen en cuenta el riesgo competitivo. Nosotros usamos dos métodos: (1) método de pseudovalores (análisis principal) y (2) método de subdistribución de hazards (análisis de sensibilidad). Los resultados de ambos fueron similares.
-   **Medidas numéricas de calibración**:
    -   Razón de eventos acumulados observados entre esperados (Razón O/E)
    -   Calibration Slope y Calibration intercept
:::

#### Métricas de calibración

En general, el modelo mostró evidencia de descalibraicón [Table 2].

##### Calibration in the large - Razón O/E

Calibration in the large es una métrica de calibración que resumen si el modelo predice bien en promedio los riesgos. Para el modelo a 2 años, el riesgo absoluto observado de falla renal fue de 1.52% en la población de 3a-3b-4 mientras que el promedio de riesgo predicho por el modelo fue menor: 0.96%. La razón de outcomes observados y esperados fue de 1.58 (95% CI 1.30 a 1.76), lo que indica que, en promedio, las predicciones subestimaron el riesgo real de falla renal. Una subestimación de riesgo ligeramente menor fue encontrada para el modelo a 2 años en la población con falla renal 3b-4 (O/E = 1.34; 95% CI 1.13 a 1.54).

Por el contrario, el riesgo observado de falla renal fue ligeramente superior pero similar al riesgo predicho promedio por el modelo a 5 años en población con estadío 3a-3b-4 (O/E = 1.06; 95% CI 0.94 a 1.19). Del mismo modo, el riesgo observado de falla renal fue ligeramente inferior pero similar al riesgo predicho promedio por el modelo a 5 años en población con estadío 3b-4 con un O/E de 0.9 (95% CI 0.76 a 1.04).

```{r}
#| label: calibracion-in-the-large

# Calibration (O/E) -------------------------------------------------------

# Seleccion del grupo: Stages 3-4----
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")))
primary_event <- 1

# A 2 años----
horizon <- 2

vdata$pred <- vdata$risk2y

# First calculate Aalen-Johansen estimate (as 'observed')
obj <- summary(survfit(Surv(time5y, status) ~ 1, 
                        data = vdata), 
                times = horizon)

aj <- list(
  "obs" = obj$pstate[, primary_event + 1],
  "se" = obj$std.err[, primary_event + 1]
)


# Calculate O/E
OE <- aj$obs / mean(vdata$pred)

# For the confidence interval we use method proposed in Debray et al. (2017) doi:10.1136/bmj.i6460
k <- 2
alpha <- 0.05
OE_summary <- cbind(
  "OE" = OE,
  "Lower .95" = exp(log(OE - qnorm(1 - alpha / 2) * aj$se / aj$obs)),
  "Upper .95" = exp(log(OE + qnorm(1 - alpha / 2) * aj$se / aj$obs))
)

OE_summary2a <- round(OE_summary, k)

avg_obs <- cbind(
  "avg_obs" = aj$obs * 100,
  "Lower .95" = 100 * (aj$obs - qnorm(1 - alpha / 2) * aj$se),
  "Upper .95" = 100 * (aj$obs + qnorm(1 - alpha / 2) * aj$se)
)

avg_pred <- cbind(
  "avg_pred" = mean(vdata$pred) * 100
)

avg_obs2a <- round(avg_obs, k)
avg_pred2a <- round(avg_pred, k)

# Seleccion del grupo: Stages 3-4----
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")))
primary_event <- 1

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y



# First calculate Aalen-Johansen estimate (as 'observed')
obj <- summary(survfit(Surv(time5y, status) ~ 1, 
                        data = vdata), 
                times = horizon)

aj <- list(
  "obs" = obj$pstate[, primary_event + 1],
  "se" = obj$std.err[, primary_event + 1]
)


# Calculate O/E
OE <- aj$obs / mean(vdata$pred)

# For the confidence interval we use method proposed in Debray et al. (2017) doi:10.1136/bmj.i6460
k <- 2
alpha <- 0.05
OE_summary <- cbind(
  "OE" = OE,
  "Lower .95" = exp(log(OE - qnorm(1 - alpha / 2) * aj$se / aj$obs)),
  "Upper .95" = exp(log(OE + qnorm(1 - alpha / 2) * aj$se / aj$obs))
)

OE_summary5a <- round(OE_summary, k)

avg_obs <- cbind(
  "avg_obs" = aj$obs * 100,
  "Lower .95" = 100 * (aj$obs - qnorm(1 - alpha / 2) * aj$se),
  "Upper .95" = 100 * (aj$obs + qnorm(1 - alpha / 2) * aj$se)
)

avg_pred <- cbind(
  "avg_pred" = mean(vdata$pred) * 100
)

avg_obs5a <- round(avg_obs, k)
avg_pred5a <- round(avg_pred, k)

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")))
primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y


# First calculate Aalen-Johansen estimate (as 'observed')
obj <- summary(survfit(Surv(time5y, status) ~ 1, 
                        data = vdata), 
                times = horizon)

aj <- list(
  "obs" = obj$pstate[, primary_event + 1],
  "se" = obj$std.err[, primary_event + 1]
)

# Calculate O/E
OE <- aj$obs / mean(vdata$pred)

# For the confidence interval we use method proposed in Debray et al. (2017) doi:10.1136/bmj.i6460
k <- 2
alpha <- 0.05
OE_summary <- cbind(
  "OE" = OE,
  "Lower .95" = exp(log(OE - qnorm(1 - alpha / 2) * aj$se / aj$obs)),
  "Upper .95" = exp(log(OE + qnorm(1 - alpha / 2) * aj$se / aj$obs))
)

OE_summary2b <- round(OE_summary, k)

avg_obs <- cbind(
  "avg_obs" = aj$obs * 100,
  "Lower .95" = 100 * (aj$obs - qnorm(1 - alpha / 2) * aj$se),
  "Upper .95" = 100 * (aj$obs + qnorm(1 - alpha / 2) * aj$se)
)

avg_pred <- cbind(
  "avg_pred" = mean(vdata$pred) * 100
)

avg_obs2b <- round(avg_obs, k)
avg_pred2b <- round(avg_pred, k)

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")))
primary_event <- 1

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y


# First calculate Aalen-Johansen estimate (as 'observed')
obj <- summary(survfit(Surv(time5y, status) ~ 1, 
                        data = vdata), 
                times = horizon)

aj <- list(
  "obs" = obj$pstate[, primary_event + 1],
  "se" = obj$std.err[, primary_event + 1]
)


# Calculate O/E
OE <- aj$obs / mean(vdata$pred)

# For the confidence interval we use method proposed in Debray et al. (2017) doi:10.1136/bmj.i6460
k <- 2
alpha <- 0.05
OE_summary <- cbind(
  "OE" = OE,
  "Lower .95" = exp(log(OE - qnorm(1 - alpha / 2) * aj$se / aj$obs)),
  "Upper .95" = exp(log(OE + qnorm(1 - alpha / 2) * aj$se / aj$obs))
)

OE_summary5b <- round(OE_summary, k)

avg_obs <- cbind(
  "avg_obs" = aj$obs * 100,
  "Lower .95" = 100 * (aj$obs - qnorm(1 - alpha / 2) * aj$se),
  "Upper .95" = 100 * (aj$obs + qnorm(1 - alpha / 2) * aj$se)
)

avg_pred <- cbind(
  "avg_pred" = mean(vdata$pred) * 100
)

avg_obs5b <- round(avg_obs, k)
avg_pred5b <- round(avg_pred, k)
```

##### Calibration intercept y slope

```{r}
#| label: calibration-intercept-slope

# Seleccion del grupo: Stages 3-4----
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")))
primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y

# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)

# Note:
# - 'pseudos' is the data.frame with ACTUAL pseudo-observations, not the smoothed ones
# - Column ID is not the id in vdata; it is just a number assigned to each row of
# the original validation data sorted by time and event indicator
pseudos$cll_pred <- log(-log(1 - pseudos$risk)) # add the cloglog risk ests

# Fit model for calibration intercept
fit_cal_int <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Fit model for calibration slope
fit_cal_slope <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)
```

La prueba conjunta de mala calibración reveló lo siguiente:

```{r}
# Perform joint test on intercept and slope
betas <- fit_cal_slope$beta
vcov_mat <- fit_cal_slope$vbeta
wald <- drop(betas %*% solve(vcov_mat) %*% betas)
pchisq(wald, df = 2, lower.tail = FALSE)
```

```{r}
k <- 2
res_cal <- rbind(

  # Value, confidence interval and test for calibration intercept
  "Intercept, Stages 3-4, 2 year" = with(
    summary(fit_cal_int)$mean,
    c(
      "estimate" = estimate,
      `2.5 %` = estimate - qnorm(0.975) * san.se,
      `97.5 %` = estimate + qnorm(0.975) * san.se
    )
  ),

  # Value, confidence interval and test for calibration slope
  "Slope, Stages 3-4, 2 year" = with(
    summary(fit_cal_slope)$mean["cll_pred", ],
    c(
      "estimate" = 1 + estimate,
      `2.5 %` = 1 + (estimate - qnorm(0.975) * san.se),
      `97.5 %` = 1 + (estimate + qnorm(0.975) * san.se)
    )
  )
)

res_cal2a <- round(res_cal, k)

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y

# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)

# Note:
# - 'pseudos' is the data.frame with ACTUAL pseudo-observations, not the smoothed ones
# - Column ID is not the id in vdata; it is just a number assigned to each row of
# the original validation data sorted by time and event indicator
pseudos$cll_pred <- log(-log(1 - pseudos$risk)) # add the cloglog risk ests

# Fit model for calibration intercept
fit_cal_int <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Fit model for calibration slope
fit_cal_slope <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)
```

La prueba conjunta de mala calibración reveló lo siguiente:

```{r}
# Perform joint test on intercept and slope
betas <- fit_cal_slope$beta
vcov_mat <- fit_cal_slope$vbeta
wald <- drop(betas %*% solve(vcov_mat) %*% betas)
pchisq(wald, df = 2, lower.tail = FALSE)
```

```{r}
k <- 2
res_cal <- rbind(

  # Value, confidence interval and test for calibration intercept
  "Intercept, Stages 3-4, 5 year" = with(
    summary(fit_cal_int)$mean,
    c(
      "estimate" = estimate,
      `2.5 %` = estimate - qnorm(0.975) * san.se,
      `97.5 %` = estimate + qnorm(0.975) * san.se
    )
  ),

  # Value, confidence interval and test for calibration slope
  "Slope, Stages 3-4, 5 year" = with(
    summary(fit_cal_slope)$mean["cll_pred", ],
    c(
      "estimate" = 1 + estimate,
      `2.5 %` = 1 + (estimate - qnorm(0.975) * san.se),
      `97.5 %` = 1 + (estimate + qnorm(0.975) * san.se)
    )
  )
)

res_cal5a <- round(res_cal, k)

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")))
primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y

# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)

# Note:
# - 'pseudos' is the data.frame with ACTUAL pseudo-observations, not the smoothed ones
# - Column ID is not the id in vdata; it is just a number assigned to each row of
# the original validation data sorted by time and event indicator
pseudos$cll_pred <- log(-log(1 - pseudos$risk)) # add the cloglog risk ests

# Fit model for calibration intercept
fit_cal_int <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Fit model for calibration slope
fit_cal_slope <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)
```

La prueba conjunta de mala calibración reveló lo siguiente:

```{r}
# Perform joint test on intercept and slope
betas <- fit_cal_slope$beta
vcov_mat <- fit_cal_slope$vbeta
wald <- drop(betas %*% solve(vcov_mat) %*% betas)
pchisq(wald, df = 2, lower.tail = FALSE)
```

```{r}
k <- 2
res_cal <- rbind(

  # Value, confidence interval and test for calibration intercept
  "Intercept, Stages 3b-4, 2 year" = with(
    summary(fit_cal_int)$mean,
    c(
      "estimate" = estimate,
      `2.5 %` = estimate - qnorm(0.975) * san.se,
      `97.5 %` = estimate + qnorm(0.975) * san.se
    )
  ),

  # Value, confidence interval and test for calibration slope
  "Slope, Stages 3b-4, 2 year" = with(
    summary(fit_cal_slope)$mean["cll_pred", ],
    c(
      "estimate" = 1 + estimate,
      `2.5 %` = 1 + (estimate - qnorm(0.975) * san.se),
      `97.5 %` = 1 + (estimate + qnorm(0.975) * san.se)
    )
  )
)

res_cal2b <- round(res_cal, k)

# A 5 años----
horizon <- 5
vdata$pred <- vdata$risk5y

# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)

# Note:
# - 'pseudos' is the data.frame with ACTUAL pseudo-observations, not the smoothed ones
# - Column ID is not the id in vdata; it is just a number assigned to each row of
# the original validation data sorted by time and event indicator
pseudos$cll_pred <- log(-log(1 - pseudos$risk)) # add the cloglog risk ests

# Fit model for calibration intercept
fit_cal_int <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Fit model for calibration slope
fit_cal_slope <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)
```

La prueba conjunta de mala calibración reveló lo siguiente:

```{r}
# Perform joint test on intercept and slope
betas <- fit_cal_slope$beta
vcov_mat <- fit_cal_slope$vbeta
wald <- drop(betas %*% solve(vcov_mat) %*% betas)
pchisq(wald, df = 2, lower.tail = FALSE)
```

```{r}
k <- 2
res_cal <- rbind(

  # Value, confidence interval and test for calibration intercept
  "Intercept, Stages 3b-4, 5 year" = with(
    summary(fit_cal_int)$mean,
    c(
      "estimate" = estimate,
      `2.5 %` = estimate - qnorm(0.975) * san.se,
      `97.5 %` = estimate + qnorm(0.975) * san.se
    )
  ),

  # Value, confidence interval and test for calibration slope
  "Slope, Stages 3b-4, 5 year" = with(
    summary(fit_cal_slope)$mean["cll_pred", ],
    c(
      "estimate" = 1 + estimate,
      `2.5 %` = 1 + (estimate - qnorm(0.975) * san.se),
      `97.5 %` = 1 + (estimate + qnorm(0.975) * san.se)
    )
  )
)

res_cal5b <- round(res_cal, k)
```

#### Curvas y cinturón de calibración usando pseudovalues y lowess

```{r}
#| label: loess-pseudo

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  theme_bw() + 
  xlab("Predicted risks") + 
  ylab("Observed outcome proportions") -> p1

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +   
  xlab("Predicted risks") + 
  ylab("Observed outcome proportions") + 
  theme_bw() -> p2

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  theme_bw() + 
  xlab("Predicted risks") + 
  ylab("Observed outcome proportions") -> p3

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +   
  xlab("Predicted risks") + 
  ylab("Observed outcome proportions") + 
  theme_bw() -> p4

p1a <- p1 + 
  labs(title = "CKD Stages 3a-3b-4\n(2 year KFRE)") + 
  theme(plot.title = element_text(hjust = 0.5))
p2a <- p2 + labs(title = "CKD Stages 3a-3b-4\n(5 year KFRE)") + 
  theme(plot.title = element_text(hjust = 0.5))
p3a <- p3 + labs(title = "CKD Stages 3b-4\n(2 year KFRE)") + 
  theme(plot.title = element_text(hjust = 0.5))
p4a <- p4 + labs(title = "CKD Stages 3b-4\n(5 year KFRE)") + 
  theme(plot.title = element_text(hjust = 0.5))

(p1a | p2a) / (p3a | p4a) + plot_annotation(tag_levels = 'A') -> plot_calibration

ggsave(filename = "Plot_Calibration.png", 
       device = "png", 
       plot = plot_calibration, 
       path = "Figures", 
       scale = 2, 
       dpi = 300)
```

```{r}
#| label: fig-loess-pseudo-ggpplot
#| fig.width: 7
#| fig.height: 7
#| fig.align: center
#| fig-cap: "Calibration plots for each group and prediction horizon. The predicted probability is shown on the x asis, and the observed kindey failure rate is given on the y axis"
plot_calibration
```

```{r}
#| label: loess-pseudo-spanish

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  theme_bw() + 
  xlab("Riesgos predichos") + 
  ylab("Riesgos observados del desenlace") -> p1

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +   
  xlab("Riesgos predichos") + 
  ylab("Riesgos observados del desenlace") + 
  theme_bw() -> p2

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  theme_bw() + 
  xlab("Riesgos predichos") + 
  ylab("Riesgos observados del desenlace") -> p3

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y
pred <- as.matrix(vdata$pred)

# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

pseudo_vals <- data.frame(
  obs = smooth_pseudos$fit, 
  risk = pseudos$risk, 
  se = smooth_pseudos$se, 
  df = smooth_pseudos$df
  ) %>% 
  mutate(
    ll = pmax(obs - qt(0.975, df) * se, 0), 
    ul = obs + qt(0.975, df) * se
  )

pseudo_vals %>% 
  ggplot(aes(x = risk, y = obs)) +
  geom_ribbon(aes(ymin = ll, ymax = ul), fill = "grey90") + 
  geom_abline(intercept = 0, slope = 1, colour = "red", linetype = 2) + 
  geom_line() + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +   
  xlab("Riesgos predichos") + 
  ylab("Riesgos observados del desenlace") + 
  theme_bw() -> p4

p1a <- p1 + 
  labs(title = "ERC Estadios 3a-3b-4\n(KFRE a 2 años)") + 
  theme(plot.title = element_text(hjust = 0.5))
p2a <- p2 + labs(title = "ERC Estadios 3a-3b-4\n(KFRE a 5 años)") + 
  theme(plot.title = element_text(hjust = 0.5))
p3a <- p3 + labs(title = "ERC Estadios 3b-4\n(KFRE a 2 años)") + 
  theme(plot.title = element_text(hjust = 0.5))
p4a <- p4 + labs(title = "ERC Estadios 3b-4\n(KFRE a 5 años)") + 
  theme(plot.title = element_text(hjust = 0.5))

(p1a | p2a) / (p3a | p4a) + plot_annotation(tag_levels = 'A') -> plot_calibration

ggsave(filename = "Plot_Calibration_Spanish.png", 
       device = "png", 
       plot = plot_calibration, 
       path = "Figures", 
       scale = 2, 
       dpi = 300)
```

```{r}
# library(rsample)
# library(infer)
# # We can extract predicted and observed, observed will depend on degree of smoothing (bandwidth)
# dat_pseudo <- pseudo_vals %>% 
#   select(obs, risk)
# 
# pred <- dat_pseudo$risk
# 
# # Calculate difference between predicted and observed (make sure to use all estimated risks, not just unique ones)
# diff_pseudo <- pred - dat_pseudo$obs[match(pred, dat_pseudo$risk)]
# 
# # Cargar funcion 
# source("Code/R/fit_modcal.R")
# 
# abs(diff_pseudo) %>% 
#   as_tibble() %>% 
#   fit_modcal(value = value, metric = "mean", nround = 3, nreps = 100) -> ici
# 
# abs(diff_pseudo) %>% 
#   as_tibble() %>% 
#   fit_modcal(value = value, metric = "median", nround = 3, nreps = 100) -> e50
# 
# bootP90 <- function(data, indices) median(data[indices])
# 
# bootMax <- function(data, indices) max(data[indices])
# 
# b90 <- boot(abs(diff_pseudo), bootMedian, R = 100)
# bMax <- boot(abs(diff_pseudo), bootMax, R = 100)
# 
# # Collect all numerical summary measures
# numsum_pseudo <- c(
#   "ICI" = ici,
#   setNames(quantile(abs(diff_pseudo), c(0.5, 0.9)), c("E50", "E90")),
#   "Emax" = max(abs(diff_pseudo)),
#   "Root squared bias" = sqrt(mean(diff_pseudo^2))
# )
# 
# k <- 4
# res_num <- rbind(
#   "Calibration measures - pseudo observations" = numsum_pseudo
# )
# 
# # res_num <- round(res_num, k)
# kable(res_num) |>
#   kable_styling("striped", position = "center")
```

```{r}
# library(rsample)
# library(infer)
# 
# fit_modcal <- function(data, value, metric, nround, nreps, ...) {
#   
#   data %>% 
#     specify(response = value) %>% 
#     calculate(stat = metric) %>% 
#     as.numeric() %>% 
#     round(nround) -> stat_res
#   
#   data %>% 
#     specify(response = value) %>% 
#     generate(reps = nreps, type = "bootstrap") %>% 
#     calculate(stat = metric) %>% 
#     get_confidence_interval(type = "percentile") %>% 
#     round(nround) %>% 
#     as.numeric() -> stat_ic
#   
#   paste0(stat_res, " (", stat_ic[1], " to ", stat_ic[2], ")") -> cal_res
#   return(cal_res)
# }
# 
# abs(diff_pseudo) %>% 
#   as_tibble() %>% 
#   fit_modcal(value = value, metric = "mean", nround = 3, nreps = 100)
```

```{r}
# ICI$ICI[[2]]
```

### Discriminación del modelo

En general, los modelos tuvieron buena discriminación [Table 2].

#### C-index y C/D AUC tiempo dependiente

En población con estadios 3a-3b-4, el estadístico C fue de 0.853 para discriminar falla renal a 2 años y ligeramente menor para 5 años: 0.827. Valores similares fueron reportados para la población con estadios 3b-4. Consistentemente, las áreas bajo la curva ROC tiempo-dependientes (C/D AUC) tuvieron valores similares para discriminar a 2 y 5 años.

```{r}
#| label: c-index
#| warning: false
#| message: false

B <- 100

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")), 
         status_num = eventd5y)

primary_event <- 1

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

# Validation set
C_vdata <- pec::cindex(
  object = list("CauseSpecificCox" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata
)$AppCindex$CauseSpecificCox

# Bootstraping C-index to calculate the bootstrap percentile confidence intervals
set.seed(777)

C_boot <- function(B, data) {
  cstat <- c()
  n <- nrow(data)
  for (i in 1:B) {
    ids <- sample(1:n, n, TRUE)
    data_boot <- data[ids, ]
    
    pec::cindex(
    object = list("CauseSpecificCox" = as.matrix(data_boot$pred)),
    formula = Hist(time5y, status_num) ~ 1,
    cause = primary_event,
    eval.times = horizon,
    data = data_boot
    )$AppCindex$CauseSpecificCox -> cstat[i]
  }
  return(cstat)
}
set.seed(777)
C_vdata_boot <- C_boot(B = B, data = vdata)

# Time-dependent AUC ---------

# Validation data
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

alpha <- .05
k <- 3
res_discr_csh <- matrix(c(
  ## C-index
  # Validation CSH1
  C_vdata,
  quantile(C_vdata_boot, probs = alpha / 2),
  quantile(C_vdata_boot, probs = 1 - alpha / 2),

  ## Time-dependent AUC
  # Validation CSH1
  score_vdata$AUC$score$AUC,
  score_vdata$AUC$score$AUC - qnorm(1 - alpha / 2) * score_vdata$AUC$score$se,
  score_vdata$AUC$score$AUC + qnorm(1 - alpha / 2) * score_vdata$AUC$score$se
  ),
  nrow = 2, ncol = 3, byrow = T,
  dimnames = list(
  c("C-index, Stages 3b-4, 2 year", "Time dependent AUC, Stages 3b-4, 2 year"),
  c("Estimate", "Lower .95", "Upper .95")
  )
  )

res_discr_csh2b <- round(res_discr_csh, k)

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")), 
         status_num = eventd5y)

primary_event <- 1

# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

# Validation set
C_vdata <- pec::cindex(
  object = list("CauseSpecificCox" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata
)$AppCindex$CauseSpecificCox

# Bootstraping C-index to calculate the bootstrap percentile confidence intervals
set.seed(777)

C_boot <- function(B, data) {
  cstat <- c()
  n <- nrow(data)
  for (i in 1:B) {
    ids <- sample(1:n, n, TRUE)
    data_boot <- data[ids, ]
    
    pec::cindex(
    object = list("CauseSpecificCox" = as.matrix(data_boot$pred)),
    formula = Hist(time5y, status_num) ~ 1,
    cause = primary_event,
    eval.times = horizon,
    data = data_boot
    )$AppCindex$CauseSpecificCox -> cstat[i]
  }
  return(cstat)
}
set.seed(777)
C_vdata_boot <- C_boot(B = B, data = vdata)

# Time-dependent AUC ---------

# Validation data
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

alpha <- .05
k <- 3
res_discr_csh <- matrix(c(
  ## C-index
  # Validation CSH1
  C_vdata,
  quantile(C_vdata_boot, probs = alpha / 2),
  quantile(C_vdata_boot, probs = 1 - alpha / 2),

  ## Time-dependent AUC
  # Validation CSH1
  score_vdata$AUC$score$AUC,
  score_vdata$AUC$score$AUC - qnorm(1 - alpha / 2) * score_vdata$AUC$score$se,
  score_vdata$AUC$score$AUC + qnorm(1 - alpha / 2) * score_vdata$AUC$score$se
  ),
  nrow = 2, ncol = 3, byrow = T,
  dimnames = list(
  c("C-index, Stages 3b-4, 5 year", "Time dependent AUC, Stages 3b-4, 5 year"),
  c("Estimate", "Lower .95", "Upper .95")
  )
  )

res_discr_csh5b <- round(res_discr_csh, k)

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")), 
         status_num = eventd5y)

primary_event <- 1

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

# Validation set
C_vdata <- pec::cindex(
  object = list("CauseSpecificCox" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata
)$AppCindex$CauseSpecificCox

# Bootstraping C-index to calculate the bootstrap percentile confidence intervals
set.seed(777)

C_boot <- function(B, data) {
  cstat <- c()
  n <- nrow(data)
  for (i in 1:B) {
    ids <- sample(1:n, n, TRUE)
    data_boot <- data[ids, ]
    
    pec::cindex(
    object = list("CauseSpecificCox" = as.matrix(data_boot$pred)),
    formula = Hist(time5y, status_num) ~ 1,
    cause = primary_event,
    eval.times = horizon,
    data = data_boot
    )$AppCindex$CauseSpecificCox -> cstat[i]
  }
  return(cstat)
}

set.seed(777)
C_vdata_boot <- C_boot(B = B, data = vdata)

# Time-dependent AUC ---------

# Validation data
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

alpha <- .05
k <- 3
res_discr_csh <- matrix(c(
  ## C-index
  # Validation CSH1
  C_vdata,
  quantile(C_vdata_boot, probs = alpha / 2),
  quantile(C_vdata_boot, probs = 1 - alpha / 2),

  ## Time-dependent AUC
  # Validation CSH1
  score_vdata$AUC$score$AUC,
  score_vdata$AUC$score$AUC - qnorm(1 - alpha / 2) * score_vdata$AUC$score$se,
  score_vdata$AUC$score$AUC + qnorm(1 - alpha / 2) * score_vdata$AUC$score$se
  ),
  nrow = 2, ncol = 3, byrow = T,
  dimnames = list(
  c("C-index, Stages 3-4, 2 year", "Time dependent AUC, Stages 3-4, 2 year"),
  c("Estimate", "Lower .95", "Upper .95")
  )
  )

res_discr_csh2a <- round(res_discr_csh, k)

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na() %>%  
  mutate(status = factor(eventd5y, 
                         levels = c(0, 1, 2), 
                         labels = c("Alive w/o Kidney Failure", 
                                    "Kidney Failure", 
                                    "Death w/o Kidney Failure")), 
         status_num = eventd5y)

primary_event <- 1

# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

# Validation set
C_vdata <- pec::cindex(
  object = list("CauseSpecificCox" = pred),
  formula = Hist(time5y, eventd5y) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata
)$AppCindex$CauseSpecificCox

# Bootstraping C-index to calculate the bootstrap percentile confidence intervals

set.seed(777)

C_boot <- function(B, data) {
  cstat <- c()
  n <- nrow(data)
  for (i in 1:B) {
    ids <- sample(1:n, n, TRUE)
    data_boot <- data[ids, ]
    
    pec::cindex(
    object = list("CauseSpecificCox" = as.matrix(data_boot$pred)),
    formula = Hist(time5y, status_num) ~ 1,
    cause = primary_event,
    eval.times = horizon,
    data = data_boot
    )$AppCindex$CauseSpecificCox -> cstat[i]
  }
  return(cstat)
}

set.seed(777)
C_vdata_boot <- C_boot(B = B, data = vdata)

# Time-dependent AUC ---------

# Validation data
score_vdata <- Score(
  list("csh_validation" = pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

alpha <- .05
k <- 3
res_discr_csh <- matrix(c(
  ## C-index
  # Validation CSH1
  C_vdata,
  quantile(C_vdata_boot, probs = alpha / 2),
  quantile(C_vdata_boot, probs = 1 - alpha / 2),

  ## Time-dependent AUC
  # Validation CSH1
  score_vdata$AUC$score$AUC,
  score_vdata$AUC$score$AUC - qnorm(1 - alpha / 2) * score_vdata$AUC$score$se,
  score_vdata$AUC$score$AUC + qnorm(1 - alpha / 2) * score_vdata$AUC$score$se
  ),
  nrow = 2, ncol = 3, byrow = T,
  dimnames = list(
  c("C-index, Stages 3-4, 5 year", "Time dependent AUC, Stages 3-4, 5 year"),
  c("Estimate", "Lower .95", "Upper .95")
  )
  )

res_discr_csh5a <- round(res_discr_csh, k)
```

#### D statistic y R<sup>2</sup><sub>D</sub> de Royston-Sauerbrei

El estadístico D y su versión transformada para mejorar interpretabilidad: $R^2_D$ indicaron buena discriminación. Especialmente, el $R^2_D$ puede interpretarse como la varianza explicada del modelo. En todos los horizontes temporales (2 años y 5 años) y subpoblaciones (3a-3b-4 y 3b-4) se apreció que el modelo explicaba más del 53% de la varibabilida de los datos de validación externa.

```{r}
#| label: royston-d2

# function to calculate D and R2D
source("Code/R/roystonD.R")

# Seleccion del grupo: Stages 3-4----
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time2y, time5y, eventd2y, eventd5y) %>%  
  drop_na() 

primary_event <- 1

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

# Rank based on predicted values at 2 years
vdata$rank <- rank(vdata$pred,
  na.last = TRUE,
  ties.method = "first"
)

res_vdata <- royston_R2D_cmprsk(
  pred = vdata$pred,
  data = vdata,
  primary_event = 1,
  time = vdata$time2y,
  status = vdata$eventd2y
)

k <- 2
res_D <- rbind(
  "Stages 3-4, 2 year" = res_vdata
  )

res_D2a <- round(res_D, k)

# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

# Rank based on predicted values at 5 years
vdata$rank <- rank(vdata$pred,
  na.last = TRUE,
  ties.method = "first"
)

res_vdata <- royston_R2D_cmprsk(
  pred = vdata$pred,
  data = vdata,
  primary_event = 1,
  time = vdata$time5y,
  status = vdata$eventd5y
)

k <- 2
res_D <- rbind(
  "Stages 3-4, 5 year" = res_vdata
  )

res_D5a <- round(res_D, k)

 # Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y, time2y, eventd2y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

# Rank based on predicted values at 2 years

vdata$rank <- rank(vdata$pred,
  na.last = TRUE,
  ties.method = "first"
)

res_vdata <- royston_R2D_cmprsk(
  pred = vdata$pred,
  data = vdata,
  primary_event = 1,
  time = vdata$time2y,
  status = vdata$eventd2y
)

k <- 2
res_D <- rbind(
  "Stages 3b-4, 2 year" = res_vdata
  )

res_D2b <- round(res_D, k)

# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

# Rank based on predicted values at 5 years

vdata$rank <- rank(vdata$pred,
  na.last = TRUE,
  ties.method = "first"
)

res_vdata <- royston_R2D_cmprsk(
  pred = vdata$pred,
  data = vdata,
  primary_event = 1,
  time = vdata$time5y,
  status = vdata$eventd5y
)

k <- 2
res_D <- rbind(
  "Stages 3b-4, 5 year" = res_vdata
  )

res_D5b <- round(res_D, k)
```

### Rendimiento global del modelo

```{r}
#| label: overall-error
#| output: false

# Seleccion del grupo: Stages 3-4----
nboot <- 100
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time2y, time5y, eventd2y, eventd5y) %>%  
  mutate(status_num = eventd5y) %>% 
  drop_na() 

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

primary_event <- 1

# Validation data
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Bootstrapping data
score_boot <- function(B, data) {
  n <- nrow(data)
  ipa <- c()
  for (i in 1:B) {
    ids <- sample(1:n, n, T)
    data_boot <- data[ids, ]
    Score(
    list("csh_validation" = data_boot$pred),
    formula = Hist(time5y, status_num) ~ 1,
    cens.model = "km",
    data = data_boot,
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc", "brier"),
    summary = c("ipa"),
    cause = primary_event,
    plots = "calibration"
    ) -> scor
    ipa[i] <- scor[["Brier"]][["score"]][["IPA"]][[2]]
    cat(".")
  }
  return(ipa)
}

set.seed(12314)
score_boot(B = nboot, data = vdata) -> vboot

# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov_csh <- matrix(unlist(c(

  # Brier score - external validation
  score_vdata$Brier$score[model == "csh_validation"]$Brier,
  score_vdata$Brier$score[model == "csh_validation"]$lower,
  score_vdata$Brier$score[model == "csh_validation"]$upper,

  # IPA - external validation
  score_vdata$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot, probs = alpha / 2),
  quantile(vboot, probs = 1 - alpha / 2)
  )),
  nrow = 2, ncol = 3,
  byrow = T,
  dimnames =
  list(
    c("Brier, Stages 3-4, 2 year", "Scaled Brier, Stages 3-4, 2 year"),
    c("Estimate", "Lower .95 ", "Upper .95")
    )
  )

res_ov2a <- round(res_ov_csh, 3) # Digits

# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

primary_event <- 1

# Validation data
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Bootstrapping data
score_boot <- function(B, data) {
  n <- nrow(data)
  ipa <- c()
  for (i in 1:B) {
    ids <- sample(1:n, n, T)
    data_boot <- data[ids, ]
    Score(
    list("csh_validation" = data_boot$pred),
    formula = Hist(time5y, status_num) ~ 1,
    cens.model = "km",
    data = data_boot,
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc", "brier"),
    summary = c("ipa"),
    cause = primary_event,
    plots = "calibration"
    ) -> scor
    ipa[i] <- scor[["Brier"]][["score"]][["IPA"]][[2]]
    cat(".")
  }
  return(ipa)
}

set.seed(12314)
score_boot(B = nboot, data = vdata) -> vboot

# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov_csh <- matrix(unlist(c(

  # Brier score - external validation
  score_vdata$Brier$score[model == "csh_validation"]$Brier,
  score_vdata$Brier$score[model == "csh_validation"]$lower,
  score_vdata$Brier$score[model == "csh_validation"]$upper,

  # IPA - external validation
  score_vdata$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot, probs = alpha / 2),
  quantile(vboot, probs = 1 - alpha / 2)
  )),
  nrow = 2, ncol = 3,
  byrow = T,
  dimnames =
  list(
    c("Brier, Stages 3-4, 5 year", "Scaled Brier, Stages 3-4, 5 year"),
    c("Estimate", "Lower .95 ", "Upper .95")
    )
  )

res_ov5a <- round(res_ov_csh, 3) # Digits

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time2y, time5y, eventd2y, eventd5y) %>%  
  mutate(status_num = eventd5y) %>% 
  drop_na() 

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

primary_event <- 1

# Validation data
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Bootstrapping data
score_boot <- function(B, data) {
  n <- nrow(data)
  ipa <- c()
  for (i in 1:B) {
    ids <- sample(1:n, n, T)
    data_boot <- data[ids, ]
    Score(
    list("csh_validation" = data_boot$pred),
    formula = Hist(time5y, status_num) ~ 1,
    cens.model = "km",
    data = data_boot,
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc", "brier"),
    summary = c("ipa"),
    cause = primary_event,
    plots = "calibration"
    ) -> scor
    ipa[i] <- scor[["Brier"]][["score"]][["IPA"]][[2]]
    cat(".")
  }
  return(ipa)
}

set.seed(12314)
score_boot(B = nboot, data = vdata) -> vboot

# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov_csh <- matrix(unlist(c(

  # Brier score - external validation
  score_vdata$Brier$score[model == "csh_validation"]$Brier,
  score_vdata$Brier$score[model == "csh_validation"]$lower,
  score_vdata$Brier$score[model == "csh_validation"]$upper,

  # IPA - external validation
  score_vdata$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot, probs = alpha / 2),
  quantile(vboot, probs = 1 - alpha / 2)
  )),
  nrow = 2, ncol = 3,
  byrow = T,
  dimnames =
  list(
    c("Brier, Stages 3b-4, 2 year", "Scaled Brier, Stages 3b-4, 2 year"),
    c("Estimate", "Lower .95 ", "Upper .95")
    )
  )

res_ov2b <- round(res_ov_csh, 3) # Digits

# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

primary_event <- 1

# Validation data
score_vdata <- Score(
  list("csh_validation" = vdata$pred),
  formula = Hist(time5y, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Bootstrapping data
score_boot <- function(B, data) {
  n <- nrow(data)
  ipa <- c()
  for (i in 1:B) {
    ids <- sample(1:n, n, T)
    data_boot <- data[ids, ]
    Score(
    list("csh_validation" = data_boot$pred),
    formula = Hist(time5y, status_num) ~ 1,
    cens.model = "km",
    data = data_boot,
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc", "brier"),
    summary = c("ipa"),
    cause = primary_event,
    plots = "calibration"
    ) -> scor
    ipa[i] <- scor[["Brier"]][["score"]][["IPA"]][[2]]
    cat(".")
  }
  return(ipa)
}

set.seed(12314)
score_boot(B = nboot, data = vdata) -> vboot

# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov_csh <- matrix(unlist(c(

  # Brier score - external validation
  score_vdata$Brier$score[model == "csh_validation"]$Brier,
  score_vdata$Brier$score[model == "csh_validation"]$lower,
  score_vdata$Brier$score[model == "csh_validation"]$upper,

  # IPA - external validation
  score_vdata$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot, probs = alpha / 2),
  quantile(vboot, probs = 1 - alpha / 2)
  )),
  nrow = 2, ncol = 3,
  byrow = T,
  dimnames =
  list(
    c("Brier, Stages 3b-4, 5 year", "Scaled Brier, Stages 3b-4, 5 year"),
    c("Estimate", "Lower .95 ", "Upper .95")
    )
  )

res_ov5b <- round(res_ov_csh, 3) # Digits
```

```{r}
#| label: table-performance-final
#| output: false

# Average predicted risk
avg_pred <- cbind("metrica" = "Average predicted risk", 
                 avg_pred2a, 
                 avg_pred5a, 
                 avg_pred2b, 
                 avg_pred5b)

colnames(avg_pred) <- c("metrica", "est2ya", "est5ya", "est2yb", "est5yb")

avg_pred %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{est2ya}%")), 
    est5ya = as.character(glue("{est5ya}%")),
    est2yb = as.character(glue("{est2yb}%")),
    est5yb = as.character(glue("{est5yb}%"))   
  ) -> avg_pred; avg_pred

# Average observed proportion
avg_obs <- cbind("metrica" = "Average observed proportion (95% CI)", 
                 avg_obs2a, 
                 avg_obs5a, 
                 avg_obs2b, 
                 avg_obs5b)

colnames(avg_obs) <- c("metrica", "OE2a", "ll2a", "ul2a", 
                          "OE5a", "ll5a", "ul5a", 
                          "OE2b", "ll2b", "ul2b", 
                          "OE5b", "ll5b", "ul5b")
avg_obs %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{OE2a}% ({ll2a}% to {ul2a}%)")), 
    est5ya = as.character(glue("{OE5a}% ({ll5a}% to {ul5a}%)")),
    est2yb = as.character(glue("{OE2b}% ({ll2b}% to {ul2b}%)")),
    est5yb = as.character(glue("{OE5b}% ({ll5b}% to {ul5b}%)"))   
  ) %>% 
  select(metrica, starts_with("est")) -> avg_obs; avg_obs

# OE summary
OE_summary <- cbind("metrica" = "O/E ratio (95% CI)", 
                    OE_summary2a, 
                    OE_summary5a, 
                    OE_summary2b, 
                    OE_summary5b); OE_summary

colnames(OE_summary) <- c("metrica", "OE2a", "ll2a", "ul2a", 
                         "OE5a", "ll5a", "ul5a", 
                         "OE2b", "ll2b", "ul2b", 
                         "OE5b", "ll5b", "ul5b")

OE_summary %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{OE2a} ({ll2a} to {ul2a})")), 
    est5ya = as.character(glue("{OE5a} ({ll5a} to {ul5a})")),
    est2yb = as.character(glue("{OE2b} ({ll2b} to {ul2b})")),
    est5yb = as.character(glue("{OE5b} ({ll5b} to {ul5b})"))   
    ) %>% 
  select(metrica, starts_with("est")) -> OE_summary; OE_summary

# Calibration slope e intercept
res_cal <- cbind("metrica" = c("Calibration intercept (95% CI)", " Calibration slope (95% CI)"), 
                 res_cal2a, 
                 res_cal5a, 
                 res_cal2b, 
                 res_cal5b); res_cal

colnames(res_cal) <- c("metrica", "OE2a", "ll2a", "ul2a", 
                       "OE5a", "ll5a", "ul5a", 
                       "OE2b", "ll2b", "ul2b", 
                       "OE5b", "ll5b", "ul5b")
res_cal <- 
  res_cal %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{OE2a} ({ll2a} to {ul2a})")), 
    est5ya = as.character(glue("{OE5a} ({ll5a} to {ul5a})")),
    est2yb = as.character(glue("{OE2b} ({ll2b} to {ul2b})")),
    est5yb = as.character(glue("{OE5b} ({ll5b} to {ul5b})"))   
  ) %>% 
  select(metrica, starts_with("est")) 

rbind(avg_pred, avg_obs, OE_summary, res_cal) -> table_performance; table_performance

# C-index y C/D AUC
res_discr_csh <- cbind("metrica" = c("C-index up to t-years (95% CI)", " C/D AUC, at t years (95% CI)"), 
                 res_discr_csh2a, 
                 res_discr_csh5a, 
                 res_discr_csh2b, 
                 res_discr_csh5b); res_discr_csh

colnames(res_discr_csh) <- c("metrica", "OE2a", "ll2a", "ul2a", 
                       "OE5a", "ll5a", "ul5a", 
                       "OE2b", "ll2b", "ul2b", 
                       "OE5b", "ll5b", "ul5b")
res_discr_csh <- 
  res_discr_csh %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{OE2a} ({ll2a} to {ul2a})")), 
    est5ya = as.character(glue("{OE5a} ({ll5a} to {ul5a})")),
    est2yb = as.character(glue("{OE2b} ({ll2b} to {ul2b})")),
    est5yb = as.character(glue("{OE5b} ({ll5b} to {ul5b})"))   
  ) %>% 
  select(metrica, starts_with("est")) 

# D-statistic y R2D
res_D <- cbind("metrica" = "D statistic (95% CI)", 
               as.matrix(t(res_D2a[, -4])), 
               as.matrix(t(res_D5a[, -4])), 
               as.matrix(t(res_D2b[, -4])), 
               as.matrix(t(res_D5b[, -4]))); res_D

colnames(res_D) <- c("metrica", "OE2a", "ll2a", "ul2a",
                             "OE5a", "ll5a", "ul5a",  
                             "OE2b", "ll2b", "ul2b",
                             "OE5b", "ll5b", "ul5b")
res_D <- 
  res_D %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{OE2a} ({ll2a} to {ul2a})")), 
    est5ya = as.character(glue("{OE5a} ({ll5a} to {ul5a})")),
    est2yb = as.character(glue("{OE2b} ({ll2b} to {ul2b})")),
    est5yb = as.character(glue("{OE5b} ({ll5b} to {ul5b})"))   
  ) %>% 
  select(metrica, starts_with("est")) 

res_D2 <- cbind("metrica" = "R2D", 
                res_D2a[4] * 100, 
                res_D5a[4] * 100, 
                res_D2b[4] * 100, 
                res_D5b[4] * 100)

colnames(res_D2) <- c("metrica", "est2ya", "est5ya", "est2yb", "est5yb")

res_D2 %>% 
  as_tibble() %>% 
  mutate(
    est2ya = as.character(glue("{est2ya}%")), 
    est5ya = as.character(glue("{est5ya}%")),
    est2yb = as.character(glue("{est2yb}%")),
    est5yb = as.character(glue("{est5yb}%"))   
  ) -> res_D2; res_D2

# Brier score
res_ov <- cbind("metrica" = c("Brier score (95% CI)", "Scaled Brier score (95% CI)"), 
                res_ov2a * c(1, 100), 
                res_ov5a * c(1, 100), 
                res_ov2b * c(1, 100), 
                res_ov5b * c(1, 100)); res_ov

colnames(res_ov) <- c("metrica", "OE2a", "ll2a", "ul2a", 
                             "OE5a", "ll5a", "ul5a", 
                             "OE2b", "ll2b", "ul2b", 
                             "OE5b", "ll5b", "ul5b")
res_ov1 <- 
  res_ov %>% 
  as_tibble() %>% 
  filter(metrica == "Brier score (95% CI)") %>% 
  mutate(
    est2ya = as.character(glue("{OE2a} ({ll2a} to {ul2a})")), 
    est5ya = as.character(glue("{OE5a} ({ll5a} to {ul5a})")),
    est2yb = as.character(glue("{OE2b} ({ll2b} to {ul2b})")),
    est5yb = as.character(glue("{OE5b} ({ll5b} to {ul5b})"))   
  ) %>% 
  select(metrica, starts_with("est"))  

res_ov2 <- 
  res_ov %>% 
  as_tibble() %>% 
  filter(metrica == "Scaled Brier score (95% CI)") %>% 
  mutate(
    est2ya = as.character(glue("{OE2a}% ({ll2a}% to {ul2a}%)")), 
    est5ya = as.character(glue("{OE5a}% ({ll5a}% to {ul5a}%)")),
    est2yb = as.character(glue("{OE2b}% ({ll2b}% to {ul2b}%)")),
    est5yb = as.character(glue("{OE5b}% ({ll5b}% to {ul5b}%)"))   
  ) %>% 
  select(metrica, starts_with("est"))  

res_ov <- rbind(res_ov1, res_ov2); res_ov


rbind(avg_pred, avg_obs, OE_summary, res_cal, res_discr_csh, res_D, res_D2, 
      res_ov) -> table_performance; table_performance

table_performance %>% 
  mutate(grupo = c(rep("Calibration", 5), rep("Discrimination", 4), rep("Overall prediction error", 2))) %>% 
  relocate(grupo, .before = "metrica") %>%
  bind_rows() %>% 
  as_grouped_data(groups = "grupo") %>% 
  flextable::as_flextable(hide_grouplabel = TRUE)  %>%
  set_header_labels(
    metrica = "Validation aspect and performance measure", 
    est2ya = "t = 2 year", 
    est5ya = "t = 5 year", 
    est2yb = "t = 2 year", 
    est5yb = "t = 5 year" 
  ) %>% 
  add_header_row(
    values = c("Validation aspect and performance measure", "CKD Stages 3a-3b-4", "CKD Stages 3b-4"), 
    colwidths = c(1, 2, 2)
  ) %>% 
  merge_v(j = 1, part = "header") %>% 
  bold(i = c(1, 7, 12)) %>% 
  autofit()  %>% 
  set_caption("Table 2. Performance measures of KFRE in the external dataset of patients with CKD Stages 3a-4 and 3b-4") %>% 
  theme_booktabs() %>%  
  bold(bold = TRUE, part = "header")  -> table_perf_final

table_perf_final %>% 
  flextable::save_as_docx(path = "Tables/Table2.docx")
```

### Table 2

```{r}
#| label: table-perf-final
table_perf_final
```

## Evaluación de utilidad del modelo predictivo

```{r}
#| label: decision-curve
#| fig.align: center
#| fig-cap: "Decision curve analysis"

# Seleccion del grupo: Stages 3-4----
group <- c("Stages 3-4")

data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time2y, time5y, eventd2y, eventd5y) %>%  
  mutate(status_num = eventd5y) %>% 
  drop_na() 

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

primary_event <- 1

vdata %>% 
  mutate(
    eventd5y = factor(eventd5y)
  ) -> vdata_dca

dca(
  Surv(time5y, eventd5y) ~ pred, 
  data = vdata_dca, 
  time = horizon, 
  thresholds = seq(0, 0.5, 0.01), 
  label = list(pred = "Prob. of kidney failure within 2 years")
  ) -> dca1 

dca1 %>% 
  plot(smooth = TRUE)  + 
  theme_bw() + 
  theme(legend.position = "bottom") -> p1


# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

primary_event <- 1

vdata %>% 
  mutate(
    eventd5y = factor(eventd5y)
  ) -> vdata_dca

dca(
  Surv(time5y, eventd5y) ~ pred, 
  data = vdata_dca, 
  time = horizon, 
  thresholds = seq(0, 0.5, 0.01), 
  label = list(pred = "Prob. of kidney failure within 5 years")
  ) -> dca2 

dca2 %>% 
  plot(smooth = TRUE) + 
  theme_bw() + 
  theme(legend.position = "bottom") -> p2

# Seleccion del grupo: Stages 3b-4----
group <- c("Stages 3b-4")

data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time2y, time5y, eventd2y, eventd5y) %>%  
  mutate(status_num = eventd5y) %>% 
  drop_na() 

# A 2 años----
vdata$pred <- vdata$risk2y
horizon <- 2 
pred <- as.matrix(vdata$pred)

primary_event <- 1

vdata %>% 
  mutate(
    eventd5y = factor(eventd5y)
  ) -> vdata_dca

dca(
  Surv(time5y, eventd5y) ~ pred, 
  data = vdata_dca, 
  time = horizon, 
  thresholds = seq(0, 0.5, 0.01), 
  label = list(pred = "Prob. of kidney failure within 2 years")
  ) -> dca3 

dca3 %>% 
  plot(smooth = TRUE)  + 
  theme_bw() +
  theme(legend.position = "bottom")  -> p3


# A 5 años----
vdata$pred <- vdata$risk5y
horizon <- 5 
pred <- as.matrix(vdata$pred)

primary_event <- 1

vdata %>% 
  mutate(
    eventd5y = factor(eventd5y)
  ) -> vdata_dca

dca(
  Surv(time5y, eventd5y) ~ pred, 
  data = vdata_dca, 
  time = horizon, 
  thresholds = seq(0, 0.5, 0.01), 
  label = list(pred = "Prob. of kidney failure within 5 years")
  ) -> dca4 

dca4 %>% 
  plot(smooth = TRUE) + 
  theme_bw()  + 
  theme(legend.position = "bottom") -> p4
```

```{r}
#| label: fig-utility
#| fig.width: 10
#| fig.height: 10
#| fig.align: center
#| fig-cap: "Decision curves analysis"
p1a <- p1 + 
  labs(title = "CKD Stages 3a-3b-4") + 
  theme(plot.title = element_text(hjust = 0.5))
p3a <- p3 + labs(title = "CKD Stages 3b-4") + 
  theme(plot.title = element_text(hjust = 0.5))

(p1a / p2) | (p3a / p4) + plot_annotation(tag_levels = 'A') -> plot_decision_curve

ggsave(filename = "plot_decision_curve.png", 
       device = "png", 
       plot = plot_decision_curve, 
       path = "Figures", 
       width = 10,
       height = 10, 
       dpi = 300)

(p1 / p2) | (p3 / p4)
```

```{r}
#| echo: false
dca1 %>% 
  as_tibble() -> data_net_dca1
```

```{r}
#| eval: false
#| echo: false
dca2 %>% 
  as_tibble()
```

```{r}
#| eval: false
#| echo: false
test_consequences(
  Surv(time5y, eventd5y) ~ pred, 
  data = vdata_dca,
  statistics = c("pos_rate", "neg_rate", "test_pos_rate", "test_neg_rate", "tp_rate",
    "fp_rate", "fn_rate", "tn_rate", "ppv", "npv", "sens", "spec", "lr_pos", "lr_neg"),
  time = horizon, 
  thresholds = seq(0, 0.5, 0.01), 
  label = list(pred = "Prob. of kidney failure within 5 years")
  )
```

## Análisis de sensibilidad

### Curvas de calibración usando método de hazards subdistribucionales

Cuando se utiliza otro enfoque de análisis, el método de hazards subdistribucionales, los gráficos de calibración revelan una mayor sobreestimación del riesgo de falla renal en comparación con los gráficos obtenidos mediante el método de pseusovalores (Ver @fig-subhazard-plot).

```{r}
#| label: fig-subhazard-plot
#| fig.width: 7
#| fig.height: 14
#| fig.align: center
#| fig-cap: "Cumulative incidence curves of the observed outcome probabilities in the population study for kidney failure or deathd without kidney failure"

# Calibration plot (flexible regression approach) -------------------------

library(survival)
library(riskRegression)

# Seleccion del grupo: Stages 3-5----
group <- c("Stages 3-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y
pred <- as.matrix(vdata$pred)
vdata$cll_pred <- log(-log(1 - pred))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)
colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))
vdata_bis <- cbind.data.frame(vdata, rcs_vdata)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time5y, eventd5y)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred
)

# Calibration plot
dat_fgr2 <- dat_fgr[order(dat_fgr$pred), ]

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y
pred <- as.matrix(vdata$pred)
vdata$cll_pred <- log(-log(1 - pred))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)
colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))
vdata_bis <- cbind.data.frame(vdata, rcs_vdata)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time5y, eventd5y)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred
)

# Calibration plot
dat_fgr5 <- dat_fgr[order(dat_fgr$pred), ]

# Seleccion del grupo: Stages 4-5----
group <- c("Stages 3b-4")

# Creacion de datsets para IA a 5 y 2 años
data %>% 
  filter(ckd_stage2 == group) -> data_filt

vdata <- data_filt %>% 
  select(id, risk2y, risk5y, time5y, eventd5y) %>%  
  drop_na()

primary_event <- 1

# A 2 años----
horizon <- 2

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk2y
pred <- as.matrix(vdata$pred)
vdata$cll_pred <- log(-log(1 - pred))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)
colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))
vdata_bis <- cbind.data.frame(vdata, rcs_vdata)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time5y, eventd5y)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred
)

# Calibration plot
dat_fgr2b <- dat_fgr[order(dat_fgr$pred), ]

# A 5 años----
horizon <- 5

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- vdata$risk5y
pred <- as.matrix(vdata$pred)
vdata$cll_pred <- log(-log(1 - pred))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)
colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))
vdata_bis <- cbind.data.frame(vdata, rcs_vdata)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time5y, eventd5y)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred
)

# Calibration plot
dat_fgr5b <- dat_fgr[order(dat_fgr$pred), ]

par(xaxs = "i", yaxs = "i", las = 1, mfrow = c(2, 2))
plot(
  x = dat_fgr2$pred,
  y = dat_fgr2$obs,
  type = "l",
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot - 2 year KFRE\n(Stages 3-4)")

plot(
  x = dat_fgr5$pred,
  y = dat_fgr5$obs,
  type = "l",
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot - 5 year KFRE\n(Stages 3-4)")

plot(
  x = dat_fgr2b$pred,
  y = dat_fgr2b$obs,
  type = "l",
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot - 2 year KFRE\n(Stages 3b-4)")

plot(
  x = dat_fgr5b$pred,
  y = dat_fgr5b$obs,
  type = "l",
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot - 5 year KFRE\n(Stages 3b-4)")
```

## Tablas suplementarias

### Tabla S1 {#sec-tableS1}

```{r}
#| label: table-S1
data %>%  
  filter(ckd_stage == "Stages 3-4") %>% 
  mutate(
    primary_event = case_when(
      eventd5ylab == "Kidney Failure" ~ "Kidney failure", 
      eventd5ylab %in% c("Alive w/o Kidney Failure", "Death w/o Kidney Failure") 
      ~ "No kidney failure"), 
    primary_event = factor(primary_event, 
                           levels = c("No kidney failure", 
                                      "Kidney failure"))) %>% 
  dplyr::select(total, sex, age, acr_cat, grf_cat, crea, eGFR_mdrd, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, ckd_class, death5y, eventd5ylab,  
                primary_event) %>% 
  gtsummary::tbl_summary(
  by = "primary_event", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})", 
    "{min}, {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  bold_labels() -> tab2


data %>%  
  filter(ckd_stage == "Stages 3-4") %>% 
  mutate(
    primary_event = case_when(
      eventd2ylab == "Kidney Failure" ~ "Kidney failure", 
      eventd2ylab %in% c("Alive w/o Kidney Failure", "Death w/o Kidney Failure") 
      ~ "No kidney failure"), 
    primary_event = factor(primary_event, 
                           levels = c("No kidney failure", 
                                      "Kidney failure"))) %>% 
  dplyr::select(total, sex, age, acr_cat, grf_cat, crea, eGFR_mdrd, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, ckd_class, death2y, eventd2ylab,  
                primary_event) %>% 
  gtsummary::tbl_summary(
  by = "primary_event", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})", 
    "{min}, {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  bold_labels() -> tab1

tbl_merge(list(tab1, tab2), 
          tab_spanner = c("2-years", "5-years")) %>% 
  modify_caption("Table S1. Baseline characteristics of the study patients at CKD Stages 3a, 3b or 4 for all predictors, stratified by outcome at 2- and 5-years") -> tab


tab %>% 
  as_flex_table() %>% 
  flextable::save_as_docx(path = "Tables/TableS1.docx")

tab %>% 
  gtsummary::as_kable_extra() %>%
  kableExtra::kable_styling("striped")
```

### Tabla S2 {#sec-tableS2}

```{r}
#| label: table-S2
data %>%  
  filter(ckd_stage2 == "Stages 3b-4") %>% 
  mutate(
    primary_event = case_when(
      eventd5ylab == "Kidney Failure" ~ "Kidney failure", 
      eventd5ylab %in% c("Alive w/o Kidney Failure", "Death w/o Kidney Failure") 
      ~ "No kidney failure"), 
    primary_event = factor(primary_event, 
                           levels = c("No kidney failure", 
                                      "Kidney failure"))) %>% 
  dplyr::select(total, sex, age, acr_cat, grf_cat, crea, eGFR_mdrd, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, ckd_class, death5y, eventd5ylab,  
                primary_event) %>% 
  gtsummary::tbl_summary(
  by = "primary_event", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})", 
    "{min}, {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  bold_labels() -> tab2


data %>%  
  filter(ckd_stage == "Stages 3-4") %>% 
  mutate(
    primary_event = case_when(
      eventd2ylab == "Kidney Failure" ~ "Kidney failure", 
      eventd2ylab %in% c("Alive w/o Kidney Failure", "Death w/o Kidney Failure") 
      ~ "No kidney failure"), 
    primary_event = factor(primary_event, 
                           levels = c("No kidney failure", 
                                      "Kidney failure"))) %>% 
  dplyr::select(total, sex, age, acr_cat, grf_cat, crea, eGFR_mdrd, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, ckd_class, death2y, eventd2ylab,  
                primary_event) %>% 
  gtsummary::tbl_summary(
  by = "primary_event", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})", 
    "{min}, {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  bold_labels() -> tab1

tbl_merge(list(tab1, tab2), 
          tab_spanner = c("2-years", "5-years")) %>% 
  modify_caption("Table S1. Baseline characteristics of the study patients at CKD Stages 3b or 4 for all predictors, stratified by outcome at 2- and 5-years") -> tab

tab %>% 
  as_flex_table() %>% 
  flextable::save_as_docx(path = "Tables/TableS2.docx")

tab %>% 
  gtsummary::as_kable_extra() %>%
  kableExtra::kable_styling("striped")
```

### Tabla S3 {#sec-tableS3}

```{r}
#| label: table-S3

data %>%  
  dplyr::select(total, sex, age, acr_cat, grf_cat, ckd_class, crea, eGFR_ckdepi, acr, 
                urine_album, urine_crea, hta, dm, bmi, album35, hemog, 
                sbp_control, dbp_control, death2y, eventd2ylab, death5y, eventd5ylab) %>% 
  gtsummary::tbl_summary(
  by = grf_cat, 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25} - {p75})", 
    "{min} - {max}"
  ),
  digits = list(all_continuous() ~ c(1, 1, 1, 1), 
                all_categorical() ~ c(0, 1)), 
  missing_text = "Missing"
  ) %>% 
  # add_p() %>% 
  bold_labels() %>% 
  modify_caption("Table S3. Baseline characteristics of the study population according CKD Stages") -> tabS3

tabS3 %>% 
  as_flex_table() %>% 
  flextable::save_as_docx(path = "Tables/TableS3.docx")

tabS3 %>% 
  gtsummary::as_kable_extra() %>%
  kableExtra::kable_styling("striped")
```

### Tabla S4 {#sec-tableS4}

```{r}
#| label: table-S4-code

data %>%  
  filter(ckd_stage == "Stages 3-4") %>% 
  dplyr::select(cas) %>% 
  gtsummary::tbl_summary(
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
  digits = list(all_categorical() ~ c(0, 1)), 
  percent = "row"
  ) %>%
  bold_labels() -> tab1a

data %>%  
  filter(ckd_stage2 == "Stages 3b-4") %>% 
  dplyr::select(cas) %>% 
  gtsummary::tbl_summary(
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
  digits = list(all_categorical() ~ c(0, 1)), 
  percent = "row"
  ) %>%
  bold_labels() -> tab1b

data %>%  
  filter(ckd_stage == "Stages 3-4") %>% 
  dplyr::select(cas, eventd5ylab) %>%  
  gtsummary::tbl_summary(
  by = "eventd5ylab",   
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
  digits = list(all_categorical() ~ c(0, 1)), 
  percent = "row"
  ) %>%  
  bold_labels() -> tab2a

data %>%  
  filter(ckd_stage2 == "Stages 3b-4") %>% 
  dplyr::select(cas, eventd5ylab) %>%  
  gtsummary::tbl_summary(
  by = "eventd5ylab",   
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
  digits = list(all_categorical() ~ c(0, 1)), 
  percent = "row"
  ) %>%  
  bold_labels() -> tab2b

tbl_merge(list(tab1a, tab2a, tab1b, tab2b), 
          tab_spanner = c("Overall", "Outcome", "Overall", "Outcome")) %>% 
  modify_caption("Table S4. Baseline characteristics of the study population according CKD Stages") -> tab

tab %>% 
  as_flex_table() %>% 
  theme_booktabs() %>%  
  bold(bold = TRUE, part = "header") %>% 
  add_header_row(
    values = c("", "CKD Stages 3a-3b-4", "CKD Stages 3a-3b-4"), 
    colwidths = c(1, 4, 4)
  ) %>% 
  flextable::save_as_docx(path = "Tables/TableS3.docx")
```

```{r}
#| label: table-S4
tab %>% 
  gtsummary::as_kable_extra() %>% 
  add_header_above(c(" ", "CKD Stages 3a-3b-4" = 4, "CKD Stages 3b-4" = 4)) %>% 
  kableExtra::kable_styling("striped")
```

### Tabla S5 {#sec-tableS5}

```{r}
#| label: table-S5

kable(
  res_ci, 
  row.names = TRUE, 
  caption = "Table S5. Cumulative incidence of Renal Failure by CKD Stages"
  ) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(" " = 1, "Stages 3a-3b-4" = 3, "Stages 3b-4" = 3))
```

### Tabla S6 {#sec-tableS6}

## Extras

::: panel-tabset
## 

## Glimpse

Se observa un resumen general del conjunto de datos con sus variables:

```{r}
#| label: view-data2
#| output: true
#| echo: true

glimpse(data)
```

## Skim

Asimismo, una exploración general de las principales medidas de resumen de cada variable del conjunto de datos:

```{r}
#| label: resume-data
#| column: page
#| echo: true

skim(data)
```

## Ticket de Reprocubilidad

```{r}
#| label: repro_ticket
#| column: page
#| echo: true
sessionInfo()
```
:::
